Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2023-06-28 manuscript.Rmd
Updates (BF section)

# Model Comparison 
The symbolic distance effect postulates that response times (RTs) are slower when digits are closer to 5 (digit effects), and that RTs may be influenced by whether the target digit is smaller or greater than 5 (side effect). Using Bayesian estimation, we are not able to directly test these hypotheses. Credible intervals often contain zero, but they always also contain lots of other values that have a chance of being the "true" parameter value, making them unsuitable to test hypotheses. Therefore, our goal for the following section is to set up a Bayesian hypothesis test for the parameters of interest in our model. Specifically, we will achieve this by comparing the predictive accuracy of a null model where we set one or some parameters to zero, and and effects models where these parameters are unconstrained.

@@

## Bayes factor
The Bayes factor (BF) is a measure of how well one model predicts the data compared to another [@jeffreys1961theory; @kass1995bayes; @rouder2018teaching]. The Bayes factor is defined as the probability of the data under one model compared to the probability of the data under another model. For instance, the Bayes factor comparing a null model ($\mathcal{M}_0$) and a full model ($\mathcal{M}_{\text{F}}$) can be obtained by:

\begin{equation}
\text{BF}_{0\text{F}} = \frac{P(\bm{Y} | \mathcal{M}_0)}{P(\bm{Y} | \mathcal{M}_{\text{F}})}. 
\end{equation}

The Bayes factor can also be interpreted as the relative evidence of two competing models. Bayes factors have a few very convenient characteristics. One of them is that Bayes factors are transitive, that is, evidence for the full model over the null model can be obtained by flipping numerator and denominator in equation 13. For a full overview of the advantages of Bayesian inference with Bayes factor, see Wagenmakers et al. [-@wagenmakers2018bayesian].

Our models consist of several parameters. Therefore, computing the Bayes factor of, for example, the normal model versus the log-normal model becomes somewhat difficult. Let the vector of all parameters of the normal model be $\bm{\theta_\text{n}}$ and the vector of all parameters for the log-normal model be $\bm{\theta_{\text{l}}}$. With the prior distributions for the two models, $f(\bm{\theta_\text{n}})$ and $f(\bm{\theta_\text{l}})$, the Bayes factor can be calculated by the following ratio of marginal likelihoods:

\begin{equation}
\text{BF}_{\text{nl}} = \frac{P(\bm{Y} | \mathcal{M}_n)}{P(\bm{Y} | \mathcal{M}_{\text{l}})} = \frac{\int_{-\infty}^{\infty}P(\bm{Y} | \mathcal{M}_\text{n}, \bm{\theta_{\text{n}}})f(\bm{\theta_{\text{n}}})d\bm{\theta_{\text{n}}}}{\int_{-\infty}^{\infty}P(\bm{Y} | \mathcal{M}_\text{l}, \bm{\theta_{\text{l}}})f(\bm{\theta_{\text{l}}})d\bm{\theta_{\text{l}}}}. 
\end{equation}

Imagine this results in $\text{BF}_{\text{nl}} = 100$. This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \Revision{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 is big enough?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. [-@rouder2018theories] argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

Same as the posterior distribution, Bayes factors can rarely be computed analytically due to the multidimensional integrals involved. The difficulty to compute Bayes factors is one of the major drawbacks of Bayesian inference. However, several algorithms are available to estimate Bayes factor from posterior samples. In the following section we will present two methods to obtain Bayes factors for the symbolic distance effect. \Revision{The first method that we discuss, Savage-Dickey density ratio, tests the general effect while disregarding individual variation (similar to type III sums of squares in ANOVA models). This means that the compared models differ in only a single parameter, that is, the general effect parameter, and individual effects are present in both models. The Savage-Dickey density ratio is computationally efficient, but can only be used for tests of a single parameter value as described before. This is why we additionally introduce a second approach, bridge-sampling, that is less computationally efficient, but more versatile. In this manuscript, we will use bridge sampling to test the general effects and individual variation together, as well as to compare the predictive accuracy of the normal and log-normal model. There have been discussions on which method should be preferred without consensus} [@van2021bayes; @van2023bayes; @rouder2023principles; @singmann2023statistics]. \Revision{Therefore, we illustrate the use of the two approaches and discuss their potential.}

>>>>

# Model Comparison 
The symbolic distance effect postulates that response times (RTs) are slower when digits are closer to 5 (digit effects), and that RTs may be influenced by whether the target digit is smaller or greater than 5 (side effect). Using Bayesian estimation, we are not able to directly test these hypotheses. Credible intervals often contain zero, but they always also contain lots of other values that have a chance of being the "true" parameter value, making them unsuitable to test hypotheses. Therefore, our goal for the following section is to set up a Bayesian hypothesis test for the parameters of interest in our model. Specifically, we will achieve this by comparing the predictive accuracy of a null model where we set one or some parameters to zero, and and effects models where these parameters are unconstrained. \Revision{This approach is called Bayes factor model comparison.}

@@

## Bayes factor
\Revision{Rather than submerging us in technical details or providing a formal justification of the Bayesian approach, we provide an intuition of Bayesian model comparison. A formal justification of Bayes factors is provided by} @jeffreys1961theory, @kass1995bayes and @rouder2018teaching. \Revision{The Bayes factor (BF) is a measure of how well one model predicts the data compared to another.}

\Revision{Bayesian analysis allows for predictions of data from any model because models are fully specified with a probability distribution for the data and priors for all parameters. Let the vector of all parameters of a model be $\bm{\theta}$. With the prior distributions of these parameters, $f(\bm{\theta})$, the prediction for data from the model can be computed by integrating over all parameters:}

\begin{equation}
p(\bm{Y} | \mathcal{M}) = \int_{\bm{\Theta}} p(\bm{Y} | \mathcal{M}, \bm{\theta})f(\bm{\theta})d\bm{\theta}. 
\end{equation}

\Revision{Note that $\int_{\bm{\Theta}}$ refers to a multidimensional integral over the entire parameter space. As our models consist of many parameters, computing these predictions for any one model becomes somewhat difficult. Once the integral is computed, the resulting probability distribution of the data, $p(\bm{Y} | \mathcal{M})$ is not a function of the model parameters anymore. This marginal probability distribution is therefore also referred to as the marginal likelihood. Before observing the data, the marginal likelihood serves as prediction for possible data; once the data are observed, the function tells us how well these specific observations were predicted by the model. The Bayes factor then is the ratio of marginal likelihoods of two models, or the relative predictive accuracy of $\mathcal{M}_1$ over $\mathcal{M}_2$:}

\begin{equation}
\text{BF}_{1,2} = \frac{P(\bm{Y} | \mathcal{M}_1)}{P(\bm{Y} | \mathcal{M}_2)}. 
\end{equation}

\Revision{There is a second popular interpretation of the Bayes factor as the relative evidence of two competing models. One of the key insights from Bayes' rule is that these two things, the relative predictive accuracy of two models for the data, and the relative evidence from the data for the two models, are one and the same} [@rouder2018teaching].

Bayes factors have a few very convenient characteristics. One of them is that Bayes factors are transitive, that is, evidence for the full model over the null model can be obtained by flipping numerator and denominator in equation 13. For a full overview of the advantages of Bayesian inference with Bayes factor, see Wagenmakers and colleagues [-@wagenmakers2018bayesian].

Imagine that $\text{BF}_{1,2} = 100$. This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \Revision{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 is big enough?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. [-@rouder2018theories] argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

\Revision{Same as the posterior distribution, Bayes factors can rarely be computed analytically.} The difficulty to compute Bayes factors is one of the major drawbacks of Bayesian inference. However, several algorithms are available to estimate Bayes factor from posterior samples. In the following section we will present two methods to obtain Bayes factors for the symbolic distance effect. \Revision{The first method that we discuss, Savage-Dickey density ratio, tests the general effect while disregarding individual variation (similar to type III sums of squares in ANOVA models). This means that the compared models differ in only a single parameter, that is, the general effect parameter, and individual effects are present in both models. The Savage-Dickey density ratio is computationally efficient, but can only be used for tests of a single parameter value as described before. This is why we additionally introduce a second approach, bridge-sampling, that is less computationally efficient, but more versatile. In this manuscript, we will use bridge sampling to test the general effects and individual variation together, as well as to compare the predictive accuracy of the normal and log-normal model. There have been discussions on which method should be preferred without consensus} [@van2021bayes; @van2023bayes; @rouder2023principles; @singmann2023statistics]. \Revision{Therefore, we illustrate the use of the two approaches and discuss their potential.}

### 2023-06-28 manuscript.tex
Updates (BF section)

\textcolor{blue}{It is clear that neither of the alternative approaches yield satisfactory results by themselves. Conceptually, a hierarchical analysis provides a compromise between the aggregated and independent approach by drawing benefits from each method. The motivation for hierarchical models, as well as their general mathematical structure, has been covered by several tutorial articles} (e.g., Singmann \& Kellen, 2019; West, Welch, \& Galecki, 2022)\textcolor{blue}{, as well as text books} (e.g., Gelman \& Hill, 2006; Hox, Moerbeek, \& Van de Schoot, 2017; Nicenboim, Schad, \& Vasishth, 2023).

@@

The symbolic distance effect postulates that response times (RTs) are slower when digits are closer to 5 (digit effects), and that RTs may be influenced by whether the target digit is smaller or greater than 5 (side effect). Using Bayesian estimation, we are not able to directly test these hypotheses. Credible intervals often contain zero, but they always also contain lots of other values that have a chance of being the ``true'' parameter value, making them unsuitable to test hypotheses. Therefore, our goal for the following section is to set up a Bayesian hypothesis test for the parameters of interest in our model. Specifically, we will achieve this by comparing the predictive accuracy of a null model where we set one or some parameters to zero, and and effects models where these parameters are unconstrained.

@@

The Bayes factor (BF) is a measure of how well one model predicts the data compared to another (Jeffreys, 1961; Kass \& Raftery, 1995; Rouder \& Morey, 2018). The Bayes factor is defined as the probability of the data under one model compared to the probability of the data under another model. For instance, the Bayes factor comparing a null model (\(\mathcal{M}_0\)) and a full model (\(\mathcal{M}_{\text{F}}\)) can be obtained by:

\begin{equation}
\text{BF}_{0\text{F}} = \frac{P(\bm{Y} | \mathcal{M}_0)}{P(\bm{Y} | \mathcal{M}_{\text{F}})}. 
\end{equation}

The Bayes factor can also be interpreted as the relative evidence of two competing models. Bayes factors have a few very convenient characteristics. One of them is that Bayes factors are transitive, that is, evidence for the full model over the null model can be obtained by flipping numerator and denominator in equation 13. For a full overview of the advantages of Bayesian inference with Bayes factor, see Wagenmakers et al. (2018).

Our models consist of several parameters. Therefore, computing the Bayes factor of, for example, the normal model versus the log-normal model becomes somewhat difficult. Let the vector of all parameters of the normal model be \(\bm{\theta_\text{n}}\) and the vector of all parameters for the log-normal model be \(\bm{\theta_{\text{l}}}\). With the prior distributions for the two models, \(f(\bm{\theta_\text{n}})\) and \(f(\bm{\theta_\text{l}})\), the Bayes factor can be calculated by the following ratio of marginal likelihoods:

\begin{equation}
\text{BF}_{\text{nl}} = \frac{P(\bm{Y} | \mathcal{M}_n)}{P(\bm{Y} | \mathcal{M}_{\text{l}})} = \frac{\int_{-\infty}^{\infty}P(\bm{Y} | \mathcal{M}_\text{n}, \bm{\theta_{\text{n}}})f(\bm{\theta_{\text{n}}})d\bm{\theta_{\text{n}}}}{\int_{-\infty}^{\infty}P(\bm{Y} | \mathcal{M}_\text{l}, \bm{\theta_{\text{l}}})f(\bm{\theta_{\text{l}}})d\bm{\theta_{\text{l}}}}. 
\end{equation}

Imagine this results in \(\text{BF}_{\text{nl}} = 100\). This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \textcolor{blue}{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 is big enough?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. (2018) argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

Same as the posterior distribution, Bayes factors can rarely be computed analytically due to the multidimensional integrals involved. The difficulty to compute Bayes factors is one of the major drawbacks of Bayesian inference. However, several algorithms are available to estimate Bayes factor from posterior samples. In the following section we will present two methods to obtain Bayes factors for the symbolic distance effect. \textcolor{blue}{The first method that we discuss, Savage-Dickey density ratio, tests the general effect while disregarding individual variation (similar to type III sums of squares in ANOVA models). This means that the compared models differ in only a single parameter, that is, the general effect parameter, and individual effects are present in both models. The Savage-Dickey density ratio is computationally efficient, but can only be used for tests of a single parameter value as described before. This is why we additionally introduce a second approach, bridge-sampling, that is less computationally efficient, but more versatile. In this manuscript, we will use bridge sampling to test the general effects and individual variation together, as well as to compare the predictive accuracy of the normal and log-normal model. There have been discussions on which method should be preferred without consensus} (Doorn, Aust, Haaf, Stefan, \& Wagenmakers, 2021, 2023; Rouder, Schnuerch, Haaf, \& Morey, 2023; Singmann et al., 2023). \textcolor{blue}{Therefore, we illustrate the use of the two approaches and discuss their potential.}

>>>>

\textcolor{blue}{It is clear that neither of the alternative approaches yield satisfactory results by themselves. Conceptually, a hierarchical analysis provides a compromise between the aggregated and independent approach by drawing benefits from each method. The motivation for hierarchical models, as well as their general mathematical structure, has been covered by several (tutorial) articles} (e.g., Efron \& Morris, 1977; Singmann \& Kellen, 2019; West, Welch, \& Galecki, 2022)\textcolor{blue}{, as well as text books} (e.g., Gelman \& Hill, 2006; Hox, Moerbeek, \& Van de Schoot, 2017; Nicenboim, Schad, \& Vasishth, 2023).

@@

The symbolic distance effect postulates that response times (RTs) are slower when digits are closer to 5 (digit effects), and that RTs may be influenced by whether the target digit is smaller or greater than 5 (side effect). Using Bayesian estimation, we are not able to directly test these hypotheses. Credible intervals often contain zero, but they always also contain lots of other values that have a chance of being the ``true'' parameter value, making them unsuitable to test hypotheses. Therefore, our goal for the following section is to set up a Bayesian hypothesis test for the parameters of interest in our model. Specifically, we will achieve this by comparing the predictive accuracy of a null model where we set one or some parameters to zero, and and effects models where these parameters are unconstrained. \textcolor{blue}{This approach is called Bayes factor model comparison.}

@@

\textcolor{blue}{Rather than submerging us in technical details or providing a formal justification of the Bayesian approach, we provide an intuition of Bayesian model comparison. A formal justification of Bayes factors is provided by} Jeffreys (1961), Kass and Raftery (1995) and Rouder and Morey (2018). \textcolor{blue}{The Bayes factor (BF) is a measure of how well one model predicts the data compared to another.}

\textcolor{blue}{Bayesian analysis allows for predictions of data from any model because models are fully specified with a probability distribution for the data and priors for all parameters. Let the vector of all parameters of a model be $\bm{\theta}$. With the prior distributions of these parameters, $f(\bm{\theta})$, the prediction for data from the model can be computed by integrating over all parameters:}

\begin{equation}
p(\bm{Y} | \mathcal{M}) = \int_{\bm{\Theta}} p(\bm{Y} | \mathcal{M}, \bm{\theta})f(\bm{\theta})d\bm{\theta}. 
\end{equation}

\textcolor{blue}{Note that $\int_{\bm{\Theta}}$ refers to a multidimensional integral over the entire parameter space. As our models consist of many parameters, computing these predictions for any one model becomes somewhat difficult. Once the integral is computed, the resulting probability distribution of the data, $p(\bm{Y} | \mathcal{M})$ is not a function of the model parameters anymore. This marginal probability distribution is therefore also referred to as the marginal likelihood. Before observing the data, the marginal likelihood serves as prediction for possible data; once the data are observed, the function tells us how well these specific observations were predicted by the model. The Bayes factor then is the ratio of marginal likelihoods of two models, or the relative predictive accuracy of $\mathcal{M}_1$ over $\mathcal{M}_2$:}

\begin{equation}
\text{BF}_{1,2} = \frac{P(\bm{Y} | \mathcal{M}_1)}{P(\bm{Y} | \mathcal{M}_2)}. 
\end{equation}

\textcolor{blue}{There is a second popular interpretation of the Bayes factor as the relative evidence of two competing models. One of the key insights from Bayes' rule is that these two things, the relative predictive accuracy of two models for the data, and the relative evidence from the data for the two models, are one and the same} (Rouder \& Morey, 2018).

Bayes factors have a few very convenient characteristics. One of them is that Bayes factors are transitive, that is, evidence for the full model over the null model can be obtained by flipping numerator and denominator in equation 13. For a full overview of the advantages of Bayesian inference with Bayes factor, see Wagenmakers and colleagues (2018).

Imagine that \(\text{BF}_{1,2} = 100\). This means that the data are 100 times more likely under the normal model than under the log-normal model, clear evidence in favor of the normal model. \textcolor{blue}{But how do we know whether a Bayes factor of 100-to-1, or 10-to-1 or 3-to-1 is big enough?} As a reminder, Bayes factors are the relative predictive accuracy of two models for the observed data. Therefore, Bayes factors are odds or ratios. Odds themselves are directly interpretable without the need for decisions or cutoffs. For example, if a presidential candidate is favored 10-to-1 over another one, then these are just the odds, and it is beside the point whether these are large or not. According to this line of thought, Rouder et al. (2018) argue that researchers should take a similar approach when interpreting and reporting Bayes factors. While there exist rules of thumb to categorize Bayes factors verbally, we suggest researchers use their research context to form substantive conclusions based on Bayes factors, not arbitrary rules of thumb.

\textcolor{blue}{Same as the posterior distribution, Bayes factors can rarely be computed analytically.} The difficulty to compute Bayes factors is one of the major drawbacks of Bayesian inference. However, several algorithms are available to estimate Bayes factor from posterior samples. In the following section we will present two methods to obtain Bayes factors for the symbolic distance effect. \textcolor{blue}{The first method that we discuss, Savage-Dickey density ratio, tests the general effect while disregarding individual variation (similar to type III sums of squares in ANOVA models). This means that the compared models differ in only a single parameter, that is, the general effect parameter, and individual effects are present in both models. The Savage-Dickey density ratio is computationally efficient, but can only be used for tests of a single parameter value as described before. This is why we additionally introduce a second approach, bridge-sampling, that is less computationally efficient, but more versatile. In this manuscript, we will use bridge sampling to test the general effects and individual variation together, as well as to compare the predictive accuracy of the normal and log-normal model. There have been discussions on which method should be preferred without consensus} (Doorn, Aust, Haaf, Stefan, \& Wagenmakers, 2021, 2023; Rouder, Schnuerch, Haaf, \& Morey, 2023; Singmann et al., 2023). \textcolor{blue}{Therefore, we illustrate the use of the two approaches and discuss their potential.}
